---
title: "Co-occurrence of Adverse Childhood Experiences (ACE) and cardiometabolic health"
author: "Juan Carlos Rivillas"
date: "03/03/2023"
---

Code walkthrough of calculation of co-occurrence of ACE and association ACE and cardiometabolic health using single and ACE score 
as per the article "Adverse Childhood Experiences and Adult Cardiometabolic Health in Colombia."

This R code includes: i) Descriptive and correlation analysis ACE, ii) estimation adjusted Odd Ratios for confounders (ACE -> diseases),
and iii) sensitivity analysis. 

Exposures: Adverse Childhood Experiences (ACE)
Outcomes: cardiovascular diseases, diabetes, hypertension, and obesity.
Disaggregation: All analyses were disaggregated by sex.
Adjusted covariates: age and ethnicity

##################################################################################
#####################    PREPARE DATA ACES-DISEASES PAPER 2   ####################
##################################################################################

#Step 1: Preparing data for analysis
```{r cars}
#Load the necessary libraries for the analysis
install.packages ('rio') #Uploading a dataset
library(rio)

install.packages ('readxl')
library(readxl)

install.packages('table1') #Tools for the Analysis of Epidemiological Data
library(table1)

install.packages('epiR') 
library(epiR)

install.packages('DescTools') #Cochran Armitage Test for trends
library(DescTools)

install.packages('car')
library(car)

install.packages('magrittr') #Pipeline Operator "%>% to chain multiple operations together
library(magrittr)

install.packages('compareGroups') 
library(compareGroups)

install.packages('ggstatsplot') #Correlations matrix and plots
library(ggstatsplot)

install.packages ('networkD3')
library(networkD3)

# install.packages("devtools")
devtools::install_github("r-lib/conflicted")

install.packages("finalfit")
library(finalfit)

library(dplyr) #Logistic regression by exposure-outcome-sex

install.packages("sjPlot") #Summary of regression Models as HTML
library(sjPlot)             
library(sjmisc)
library(sjlabelled)
library(table1)
```

#We’ll use the SABE aging population survey dataset. This dataset contains the following information about 18,044 individuals:
```{r}
#load datasets
pathways_final #Dataset with 18,044 individuals
tetrachoric_correlation1
dfcase_complete #SABE subsample with 2,813 individuals (sensitivity analysis)

#view summary of dataset
summary(pathways_final)

#find total observations in dataset
nrow(pathways_final)
```
#Split data by sex (To explore sex differences)
```{r}
data_female <- pathways_final %>% filter(sex == "Female")
data_male <- pathways_final %>% filter(sex == "Male")
```

#Save two datasets males and females (Otherwise both databases are already uplodaded on GitHub)
```{r}
writexl::write_xlsx( x = data_male, path =  '/Users/macbookpro/Documents/PhD Project/Data/SABE/dataframes/data_male.xlsx' )
writexl::write_xlsx( x = data_female, path =  '/Users/macbookpro/Documents/PhD Project/Data/SABE/dataframes/data_female.xlsx' )
```

#Step 2: Missingness analysis: check names columns, number of observations, n missing values, completed rates, and type of data (caracter or numeric) in the datframe.
```{r }
skimr::skim(pathways_final)
skimr::skim(data_female)
skimr::skim(data_male)
```


##################################################################################
#####################DESCRIPTIVE ANALYSIS ACES-DISEASES PAPER 2####################
##################################################################################

#Step3: Descriptive analysis - Analysis Individual ACEs
```{r}
#Extract descriptive data
table_2 <-table1 (~age + ethnic_group + emotional_abuse + poor_food_environment + poor_childhood_health + domestic_violence + childhood_migration + early_infection + HTA + CVD + Diabetes + Obesity + aces_score + Low_childhood_sep | sex, data=pathways_final)

table_2
```


#Descriptive analysis - Analysis of co-occurence ACES (ACEs score)

#Explore co-occurring ACE using correlation matrix
```{r}
## as a default this function outputs a correlation matrix plot
ggcorrmat(
  data     = tetrachoric_correlation1,
  colors   = c("#B2182B", "white", "#4D4D4D"),
  title    = "Descriptive co-occurrence of childhood adversities ",
  subtitle = "Tetrachroric correlation using a correlalogram"
)
```

#Create a new category with four levels:
ACE 0: No exposed to any childhood adversity
ACE 1: Exposed single childhood adversity
ACE 2: Exposed to two childhood adversity
ACE 3: Exposed to three childhood adversity
ACE 4+:Exposed to more than four childhood adversity

```{r}
pathways_final$ACE_score <- pathways_final$childhood_migration.x + pathways_final$emotional_abuse.x + pathways_final$poor_health2 + pathways_final$neglected_food + pathways_final$household_violence + pathways_final$early_infection.x

#create a vector ACE score (Sum of single ACE)
pathways_final$ACE_score <- factor(pathways_final$ACE_score,
                levels = c("0","1","2","3","4","5","6"),
                labels = c("ACE 0","ACE 1", "ACE 2","ACE 3", "ACE 4+","ACE 4+", "ACE 4+"))

Desc(pathways_final$ACE_score)
```

#Check summary of results by groups of 'aces_score'
```{r}
test_compare2 <- compareGroups( data = pathways_final, 
               formula = ACE_score  ~ 
               emotional_abuse + poor_food_environment + poor_childhood_health + domestic_violence + childhood_migration + early_infection + Low_childhood_sep)
test_compare2 

test_table5 <- createTable( test_compare2, show.ci = TRUE )
test_table5
```
```

#Step 4: Descriptive analysis - ACE variables and diseases (Overall population)
#Generate tables with n, percentages by diseases and p values
```{r}
#Cardiovascular diseases by ACE
test_compare3 <- compareGroups( data = pathways_final, 
               formula = CVD  ~ 
               emotional_abuse + poor_food_environment + poor_childhood_health + domestic_violence + childhood_migration + early_infection + ACE_score + Low_childhood_sep)

test_compare3 
test_table3 <- createTable(test_compare3, show.ci = TRUE)
test_table3
```


```{r}
#Diabetes by ACE
test_compare4 <- compareGroups( data = pathways_final, 
               formula = Diabetes  ~ 
               emotional_abuse + poor_food_environment + poor_childhood_health + domestic_violence + childhood_migration + early_infection + ACE_score + Low_childhood_sep)

test_compare4 
test_table4 <- createTable( test_compare4, show.ci = TRUE )
test_table4
```


```{r}
#Hypertension by ACE
test_compare5 <- compareGroups( data = pathways_final, 
               formula = HTA  ~ 
               emotional_abuse + poor_food_environment + poor_childhood_health + domestic_violence + childhood_migration + early_infection + ACE_score + Low_childhood_sep)

test_compare5 
test_table5 <- createTable( test_compare5, show.ci = TRUE )
test_table5
```


```{r}
#Obesity by ACE
test_compare1 <- compareGroups( data = pathways_final, 
               formula = Obesity  ~ 
               emotional_abuse + poor_food_environment + poor_childhood_health + domestic_violence + childhood_migration + early_infection + ACE_score + Low_childhood_sep)

test_compare1 
test_table1 <- createTable( test_compare1, show.ci = TRUE )
test_table1
```


##################################################################################
############## LOGISTIC REGRESSION ANALYSIS ACES-DISEASES PAPER 2 ################
##################################################################################

#We use childhood adversity and adult diseases to build a logistic regression models that predicts the probability that a given individual develop cardiovascular diseases, diabetes, hypertension, or obesity in adulthood.

#Step 5: Build logistic regression models, tables and plots.

#Change to ordinary factor for the regressions in the three datasets
```{r}
#Overall population dataset
pathways_final$Obesity = as.factor(pathways_final$Obesity)
pathways_final$HTA      = as.factor(pathways_final$HTA)
pathways_final$Diabetes = as.factor(pathways_final$Diabetes)
pathways_final$CVD      = as.factor(pathways_final$CVD)

pathways_final$sex2     = as.factor(pathways_final$sex2)
pathways_final$ACEs2    = as.factor(pathways_final$ACEs2)
pathways_final$ethnic_group2   = as.factor(pathways_final$ethnic_group2)
pathways_final$aces_score   = as.factor(pathways_final$aces_score)
pathways_final$ACE_score   = as.factor(pathways_final$ACE_score)
```                         

#Change to factor for the regressions
```{r}
#Females dataset
data_female$Obesity = as.factor(data_female$Obesity)
data_female$HTA      = as.factor(data_female$HTA)
data_female$Diabetes = as.factor(data_female$Diabetes)
data_female$CVD      = as.factor(data_female$CVD)

data_female$sex2     = as.factor(data_female$sex2)
data_female$ACEs2    = as.factor(data_female$ACEs2)
data_female$ethnic_group2   = as.factor(data_female$ethnic_group2)
data_female$aces_score   = as.factor(data_female$aces_score)
data_female$ACE_score   = as.factor(data_female$ACE_score)
```                         
                         
#Change to factor for the regressions 
```{r}
#Males dataset
data_male$Obesity  = as.factor(data_male$Obesity)
data_male$HTA      = as.factor(data_male$HTA)
data_male$Diabetes = as.factor(data_male$Diabetes)
data_male$CVD      = as.factor(data_male$CVD)

data_male$sex2     = as.factor(data_male$sex2)
data_male$ACEs2    = as.factor(data_male$ACEs2)
data_male$ethnic_group2   = as.factor(data_male$ethnic_group2)
data_male$aces_score   = as.factor(data_male$aces_score)
data_male$ACE_score   = as.factor(data_male$ACE_score)
```

#To avoid re-parametizer logictis models, we change the reference category using relevel (putting always reference the wealthiest groups)  
```{r}
#Overall population dataset
pathways_final <- within (pathways_final, ACEs2 <- relevel (ACEs2, ref = '1'))
pathways_final <- within (pathways_final, sex2 <- relevel (sex2, ref = '1'))
pathways_final <- within (pathways_final, ethnic_group2 <- relevel (ethnic_group2, ref = '1'))
pathways_final <- within (pathways_final, aces_score <- relevel (aces_score, ref = '1'))
pathways_final <- within (pathways_final, ACE_score <- relevel (ACE_score, ref = 'ACE 0'))

table(pathways_final$ACEs2)
table(pathways_final$sex2)
table(pathways_final$ethnic_group2)
table(pathways_final$aces_score)
```


```{r}
#Males dataset
data_female <- within (data_female, ACEs2 <- relevel (ACEs2, ref = '1'))
data_female <- within (data_female, ethnic_group2 <- relevel (ethnic_group2, ref = '1'))
data_female <- within (data_female, aces_score <- relevel (aces_score, ref = '1'))
data_female <- within (data_female, ACE_score <- relevel (ACE_score, ref = 'ACE 0'))

table(data_male$ACEs2)
table(data_male$sex2)
table(data_male$ethnic_group2)
table(data_male$aces_score)
```


```{r}
#Females dataset
data_male <- within (data_male, ACEs2 <- relevel (ACEs2, ref = '1'))
data_male <- within (data_male, sex2 <- relevel (sex2, ref = '1'))
data_male <- within (data_male, ethnic_group2 <- relevel (ethnic_group2, ref = '1'))
data_male <- within (data_male, aces_score <- relevel (aces_score, ref = '1'))
data_male <- within (data_male, ACE_score <- relevel (ACE_score, ref = 'ACE 0'))

table(data_male$ACEs2)
table(data_male$sex2)
table(data_male$ethnic_group2)
table(data_male$aces_score)
```


#Specify explanatory variables of interest and i) generate tables with unadjusted and adjusted odd ratios and Graphics ORs

#Graphics OR for overall population (we repeat same procedures in females and males data)
```{r}
#Overall population dataset (Cardiovascular diseases)
iv <- c("ACE_score","age", "ethnic_group2")
dv <- "CVD"

pathways_final %>% or_plot(dv, iv)

table2 <- pathways_final %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```


```{r}
#Overall population dataset (Diabetes)
iv <- c("ACE_score","age", "ethnic_group2")
dv <- "Diabetes"

pathways_final %>% or_plot(dv, iv)

table2 <- pathways_final %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```


```{r}
#Overall population dataset (Hypertension)
iv <- c("ACE_score","age", "ethnic_group2")
dv <- "HTA"

pathways_final %>% or_plot(dv, iv)

table2 <- pathways_final %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```


```{r}
#Overall population dataset (Obesity)
iv <- c("ACE_score","age", "ethnic_group2")
dv <- "Obesity"

pathways_final %>% or_plot(dv, iv)

table2 <- pathways_final %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```


# Repeat same procedure in the Females dataset.

Graphics OR for females (we repeat same procedures in males data)
```{r}
# Females dataset (Cardiovascular diseases)
iv <- c("ACE_score","age", "ethnic_group2")
dv <- "CVD"

data_female %>% or_plot(dv, iv)

table2 <- data_female %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```


```{r}
# Females dataset (Diabetes)
iv <- c("ACE_score","age","ethnic_group2")
dv <- "Diabetes"

data_female %>% or_plot(dv, iv)

table2 <- data_female %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```


```{r}
# Females dataset (hypertension)
iv <- c("ACE_score","age", "ethnic_group2")
dv <- "HTA"

data_female %>% or_plot(dv, iv)

table2 <- data_female %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```


```{r}
#Females dataset (Obesity)
iv <- c("ACE_score","age", "ethnic_group2")
dv <- "Obesity"

data_female %>% or_plot(dv, iv)

table2 <- data_female %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```


#Graphics OR for males
```{r}
# Males dataset (Cardiovascular diseases)
iv <- c("ACE_score","age", "ethnic_group2")
dv <- "CVD"

data_male %>% or_plot(dv, iv)

table2 <- data_male %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```


```{r}
# Males dataset (Diabetes)
iv <- c("ACE_score","age", "ethnic_group2")
dv <- "Diabetes"

data_male %>% or_plot(dv, iv)

table2 <- data_male %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```


```{r}
# Males dataset (Hypertension)
iv <- c("ACE_score","age", "ethnic_group2")
dv <- "HTA"

data_male %>% or_plot(dv, iv)

table2 <- data_male %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```


```{r}
# Males dataset (Obesity)
iv <- c("ACE_score","age", "ethnic_group2")
dv <- "Obesity"

data_male %>% or_plot(dv, iv)

table2 <- data_male %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```


#Perform a Cochran Armitage test for trend in binomial proportions across the levels of a single variable. 
#This test is appropriate only when one variable has two levels and the other variable is ordinal. 
#The two-level variable represents the response, and the other represents an explanatory variable with ordered levels. 
#The null hypothesis is the hypothesis of no trend, which means that the binomial proportion is the same for all levels of the explanatory variable.
```{r}
tab2 <- table(pathways_final$ACE_score, pathways_final$CVD)
CochranArmitageTest(tab2)

tab1 <- table(pathways_final$ACE_score, pathways_final$Diabetes)
CochranArmitageTest(tab1)

tab3 <- table(pathways_final$ACE_score, pathways_final$HTA)
CochranArmitageTest(tab3)

tab4 <- table(pathways_final$ACE_score, pathways_final$Obesity)
CochranArmitageTest(tab4)
```


Perform a Cochran Armitage test for trend in binomial proportions across the levels of a single variable. The null hypothesis is the hypothesis of no trend, which means that the binomial proportion is the same for all levels of the explanatory variable.
```{r}
tab2 <- table(data_female$ACE_score, data_female$CVD)
CochranArmitageTest(tab2)

tab1 <- table(data_female$ACE_score, data_female$Diabetes)
CochranArmitageTest(tab1)

tab3 <- table(data_female$ACE_score, data_female$HTA)
CochranArmitageTest(tab3)

tab4 <- table(data_female$ACE_score, data_female$Obesity)
CochranArmitageTest(tab4)
```


```{r}
tab8 <- table(data_male$ACE_score, data_male$Diabetes)
CochranArmitageTest(tab8)

tab5 <- table(data_male$ACE_score, data_male$CVD)
CochranArmitageTest(tab5)

tab6 <- table(data_male$ACE_score, data_male$HTA)
CochranArmitageTest(tab6)

tab7 <- table(data_male$ACE_score, data_male$Obesity)
CochranArmitageTest(tab7)
```


#Step 6: Sensitivity analysis

#Missingness analysi in the datframe.
```{r }
skimr::skim(dfcase_complete)
```


#Descriptive analysis - Analysis Individual ACEs
```{r}
#Extract descriptive data
table_a <-table1 (~ethnic_group + emotional_abuse + poor_food_environment + poor_childhood_health + domestic_violence + childhood_migration + early_infection + HTA + CVD + Diabetes + Obesity + Low_childhood_sep | sex, data=dfcase_complete)

table_a
```


#Create a new category with four levels:
```{r}
dfcase_complete$ACE_score <- dfcase_complete$childhood_migration.x + dfcase_complete$emotional_abuse.x + dfcase_complete$poor_childhood_health.x + dfcase_complete$poor_food_environment.x + dfcase_complete$domestic_violence.x + dfcase_complete$early_infection.x

#create a vector ACE score (Sum of single ACE)
dfcase_complete$ACE_score <- factor(dfcase_complete$ACE_score,
                levels = c("0","1","2","3","4","5","6"),
                labels = c("ACE 0","ACE 1", "ACE 2","ACE 3", "ACE 4+","ACE 4+", "ACE 4+"))

Desc(dfcase_complete$ACE_score)
```


#Analysis co-occurence ACEs
```{r}
test_compare2 <- compareGroups( data = dfcase_complete, 
               formula = ACE_score  ~ 
               emotional_abuse + poor_food_environment + poor_childhood_health + domestic_violence + childhood_migration + early_infection + Low_childhood_sep)
test_compare2 

test_table5 <- createTable( test_compare2, show.ci = TRUE )
test_table5
```


#Build logistic regression models, tables and plots.
#Change to ordinary factor for the regressions in the three datasets
```{r}
#Overall population dataset
dfcase_complete$age = as.factor(dfcase_complete$age)

dfcase_complete$CVD      = as.factor(dfcase_complete$CVD)
dfcase_complete$HTA      = as.factor(dfcase_complete$HTA)
dfcase_complete$Diabetes = as.factor(dfcase_complete$Diabetes)
dfcase_complete$Obesity = as.factor(dfcase_complete$Obesity)

dfcase_complete$gender     = as.factor(dfcase_complete$gender)
dfcase_complete$ethnic_group   = as.factor(dfcase_complete$ethnic_group)
dfcase_complete$ACE_score   = as.factor(dfcase_complete$ACE_score)
```                         

#To avoid re-parametizer logictis models, we change the reference category using relevel (putting always reference the wealthiest groups)  
```{r}
#Overall population dataset
dfcase_complete <- within (dfcase_complete, gender <- relevel (gender, ref = '1'))
dfcase_complete <- within (dfcase_complete, ethnic_group <- relevel (ethnic_group, ref = 'white'))
dfcase_complete <- within (dfcase_complete, ACE_score <- relevel (ACE_score, ref = 'ACE 0'))

table(dfcase_complete$gender)
table(dfcase_complete$ethnic_group)
table(dfcase_complete$ACE_score)
```

#To generate tables with unadjusted and adjusted odd ratios and Graphics ORs
```{r}
#SABE subsample (cardiovacular diseases)
iv <- c("ACE_score","gender", "ethnic_group")
dv <- "CVD"

dfcase_complete %>% or_plot(dv, iv)

table2 <- dfcase_complete %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```


```{r}
#SABE subsample (Hypertension)
iv <- c("ACE_score","sex", "ethnic_group")
dv <- "HTA"

dfcase_complete %>% or_plot(dv, iv)

table2 <- dfcase_complete %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```


```{r}
#SABE subsample (Diabetes)
iv <- c("ACE_score","gender", "ethnic_group")
dv <- "Diabetes"

dfcase_complete %>% or_plot(dv, iv)

table2 <- dfcase_complete %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```


```{r}
#SABE subsample  (Obesity)
iv <- c("ACE_score","gender", "ethnic_group")
dv <- "Obesity"

dfcase_complete %>% or_plot(dv, iv)

table2 <- dfcase_complete %>%
finalfit(dv, iv,dependent_label_prefix = "")

table2
```


#Validation Logistic regression with sjPlot package
```{r}
m1 <- glm(CVD ~ ACE_score + gender + ethnic_group, data = dfcase_complete, family = binomial(logit))
tab_model (m1)

m2 <- glm(HTA ~ ACE_score + gender + ethnic_group, data = dfcase_complete, family = binomial(logit))
tab_model (m2)

m3 <- glm(Diabetes ~ ACE_score + gender + ethnic_group, data = dfcase_complete, family = binomial(logit))
tab_model (m3)

m4 <- glm(Obesity ~ ACE_score + gender + ethnic_group,data = dfcase_complete, family = binomial(logit))
tab_model (m4)
```

###############################################################################
########################      END     #########################################
###############################################################################



